Weighted Model Counting with Conditional Weights

### 2020-08-12 paper/paper.tex
Revision based on feedback

\begin{abstract}
  Weighted model counting (WMC) has emerged as the unifying inference mechanism
  across many (probabilistic) domains. Encoding an inference problem as an
  instance of WMC typically necessitates adding extra literals and clauses. This
  is partly so because the predominant definition of WMC assigns weights to
  models based on weights on literals, and this severely restricts what
  probability distributions can be represented. We propose a way to encode
  conditional weights on literals analogously to conditional probabilities. This
  representation can be as succinct as standard WMC with weights on literals but
  can also expand as needed to represent probability distributions with less
  structure. To demonstrate the performance benefits of conditional weights over
  the addition of extra literals, we develop a new WMC encoding for Bayesian
  networks and adapt a state-of-the-art WMC algorithm ADDMC to the new format.
  Our experiments show that the new encoding results in significantly faster
  inference on most benchmark instances. Furthermore, the same idea can
  be adapted to other WMC algorithms and other problem domains.
\end{abstract}

@@

%Many WMC algorithms have been proposed
%based primarily either on knowledge compilation
%\cite{DBLP:conf/ecai/Darwiche04,DBLP:conf/ijcai/OztokD15,DBLP:conf/ijcai/LagniezM17}
%or search \cite{DBLP:conf/aaai/SangBK05}.

Weighted model counting (WMC), i.e., an extension of model counting ($\#\SAT{}$)
that assigns a weight to every model \cite{DBLP:conf/aaai/SangBK05} has emerged
as the unifying inference mechanism across many (probabilistic) domains. It has
been applied to Bayesian networks
\cite{DBLP:conf/aaai/SangBK05,DBLP:books/daglib/0024906} and probabilistic
graphical models more generally \cite{DBLP:conf/ecsqaru/ChoiKD13}, probabilistic
logic programming languages \cite{DBLP:journals/tplp/FierensBRSGTJR15} and
probabilistic programs with discrete probability distributions
\cite{DBLP:journals/corr/abs-2005-09089}. Over the last fifteen years, WMC has
been extended and generalised in many ways, e.g., to handle continuous
probability distributions \cite{DBLP:conf/ijcai/BellePB15}, first-order
probabilistic theories
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16}, and infinite
domains \cite{DBLP:conf/aaai/Belle17}. Furthermore, the idea of weights in WMC
has been generalised from non-negative real numbers to an arbitrary semiring,
making it applicable to domains such as parsing, dynamic programming, constraint
programming, databases, and networks \cite{DBLP:journals/japll/KimmigBR17}.

@@

% TODO: highlight the main strengths

@@

@@

\paragraph{ADDs and their use in probabilistic inference.} ADDs provide an
efficient way to manipulate functions from a Boolean algebra to any algebraic
structure (most commonly the real numbers)
\cite{DBLP:journals/fmsd/BaharFGHMPS97}. They have been used to represent value
functions in Markov decision processes \cite{DBLP:conf/uai/HoeySHB99} and, for
Bayesian network inference, to represent each conditional probability table
(CPT) as an ADD \cite{DBLP:conf/icml/ZhaoMP15} as well as by combining ADDs and
arithmetic circuits into a  single representation
\cite{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous in this
situation because of their ability to fully exploit context-specific
independence, i.e., the observable structure within a CPT that is not inherited
from the structure of the Bayesian network \cite{DBLP:conf/uai/BoutilierFGK96}.

@@

Note that in this case, we chose to interpret $\nu_b$ as $\Pr(b \mid a)$
while---with a different definition of $\nu_b$ that represents the joint
probability distribution $\Pr(a, b)$---$\nu_b$ by itself could induce $\mu$. In
general, however, factorising the full weight function into several smaller
functions often results in weight functions with smaller domains which leads to
increased efficiency and decreased memory usage \cite{DBLP:conf/aaai/DudekPV20}.

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\cite{DBLP:journals/ai/ChaviraD08}. We can indeed show that this is always
possible, however, as ensuing sections will demonstrate, it often makes the
inference task much harder.

@@

In this section, we describe a way to encode Bayesian networks into WMC without
restricting oneself to factorable measures and thus having to add extra
variables. In line with the names of the previous encodings, we shall call this
encoding \texttt{db21}.

A Bayesian network is a directed acyclic graph with random variables as vertices
that defines a probability distribution over them. Let $\mathcal{V}$ denote this
set of random variables. For any random variable $X \in \mathcal{V}$, let $\im
X$ denote its set of values and $\mathrm{pa}(X)$ its set of parents. The
full probability distribution is then equal to $\prod_{X \in \mathcal{V}} \Pr(X
\mid \mathrm{pa}(X))$. For discrete Bayesian networks (and we only consider
discrete networks in this paper), each factor of this product can be represented
by a CPT. See \cref{fig:example_bn} for an example Bayesian network which we
will refer to throughout this section. For this network, $\mathcal{V} = \{ W, F,
T \}$, $\mathrm{pa}(W) = \emptyset$, $\mathrm{pa}(F) = \mathrm{pa}(T) = \{ W
\}$, $\im W = \im F = \{0, 1 \}$, and $\im T = \{ l, m, h \}$.

@@

For a binary random variable $X$, $\mathrm{CPT}_X$ is simply a sum of smaller
functions, one for each row of the CPT. If $X$ has more than two values, we also
multiply $\mathrm{CPT}_X$ by some `clause' functions that restrict the value of
$\phi(T)$ to zero whenever $|\mathcal{E}(X) \cap T| \ne 1$. For the Bayesian
network in \cref{fig:example_bn}, we get:
\begin{align*}
  \mathrm{CPT_F} &= 0.6[\lambda_{F=1}] \cdot [\lambda_{W=1}] + 0.4[\lambda_{F=0}] \cdot [\lambda_{W=1}] \\
                 &+ 0.1[\lambda_{F=1}] \cdot [\lambda_{W=0}] + 0.9[\lambda_{F=0}] \cdot [\lambda_{W=0}], \\
  \mathrm{CPT_T} &= ([\lambda_{T=l}] + [\lambda_{T=m}] + [\lambda_{T=h}]) \\
                 &\cdot (\overline{[\lambda_{T=l}]} + \overline{[\lambda_{T=m}]}) \cdot (\overline{[\lambda_{T=l}]} + \overline{[\lambda_{T=h}]}) \\
                 &\cdot (\overline{[\lambda_{T=m}]} + \overline{[\lambda_{T=h}]}) \cdot \dots.
\end{align*}

@@

\Cref{alg:encoding} produces a function with a Boolean algebra as its domain.
This function can be represented by an ADD
\cite{DBLP:journals/fmsd/BaharFGHMPS97}. The core of ADDMC works by taking an
ADD $\psi\colon 2^{U} \to \mathbb{R}_{\ge 0}$ (expressed as a product of smaller
ADDs) and returning $(\exists_U\psi)(\emptyset)$
\cite{DBLP:conf/aaai/DudekPV20}. In this section, we prove that the function
$\phi$ produced by \cref{alg:encoding} can be used by ADDMC to correctly compute
any marginal probability of the Bayesian network that was encoded as
$\phi$.\footnote{It can just as well compute \emph{any} probability expressed
  using the random variables in $\mathcal{V}$. We only focus on marginal
  probabilities for notational simplicity.} We begin with \cref{lemma:cpt} which
shows that any conditional weight function produces the right answer when given
a valid encoding of variable-value assignments relevant to the CPT.

@@

For all encodings other than \texttt{db21}, we use their implementation in
Ace~3.0\footnote{\url{http://reasoning.cs.ucla.edu/ace/}} with
\texttt{-encodeOnly} and \texttt{-noEclause} flags. However, Ace was not used to
encode evidence, as preliminary experiments revealed that the evidence-encoding
implementation contains bugs that can lead to incorrect answers or a Java
exception being thrown on some instances of the data set (and the source code is
not publicly available). Instead, we simply list all the evidence as additional
clauses in the encoding.

@@

While encoding time itself was not measured because of different languages of
implementation, it is worth noting that \texttt{db21} is linear in the total
number of CPT rows and so is unlikely to be slower than, e.g., \texttt{cd06},
which relies on solving $\NP{}$-complete problems as part of the encoding process
\cite{DBLP:conf/sat/ChaviraD06}.

@@

\begin{table}
  \centering
  \caption{The numbers of instances solved by ADDMC with each encoding
    (uniquely, faster than with others, and in total) under the described time
    and memory constraints (out of 1216 instances)}%
  \begin{tabular}{lrrr}
    \toprule
    Encoding & Unique & Fastest & Total \\
    \midrule
    \texttt{cd05} & 2 & 3 & 372 \\
    \texttt{cd06} & 0 & 1 & 351 \\
    \texttt{d02} & 41 & 99 & 726 \\
    \texttt{db21} & 228 & 871 & 901 \\
    \texttt{sbk05} & 6 & 36 & 687 \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{figure*}
  \centering
  \input{scatter.tex}%
  \caption{ADDMC inference time using \texttt{db21} compared to \texttt{d02}
    (left) and \texttt{sbk05} (right) for each instance in all data sets.}
\end{figure*}

% TODO: write. then run through a grammar checker.
\paragraph{Observations.}
\begin{itemize}
\item The order of the encodings from best to worst is exactly the opposite of
  that in the previous literature.
\item Our encoding is particularly promising on instances of Grid networks.
  They are set up so that the entire Bayesian network is relevant to the
  query.
\item However, \texttt{db21} struggles with instances from Mastermind and
  Random Blocks domains (but so does \texttt{sbk05}). We conjecture that this
  is so either because the particular structure of these problems confuse the
  heuristics used by ADDMC or because these instances allow for significant
  simplifications that are exploited by $\mathtt{d02}$ but not
  $\mathtt{db21}$.
\item An observation from the cumulative plot: \texttt{db21} solves the same
  number of instances in \SI{6.374}{\second} as \texttt{d02} does in
  \SI{1000}{\second}.
\end{itemize}

\begin{table}
  \centering
  \caption{Asymptotic upper bounds on the numbers of variables and clauses/ADDs
    for each encoding}
  \label{tbl:asymptotes}
  \begin{tabular}{lcc}
    \toprule
    Encoding(s) & Variables & Clauses/ADDs \\
    \midrule
    \texttt{cd05}, \texttt{cd06}, \texttt{sbk05} & $\mathcal{O}(nv^{d+1})$ & $\mathcal{O}(nv^{d+1})$ \\
    \texttt{d02} & $\mathcal{O}(nv^{d+1})$ & $\mathcal{O}(ndv^{d+1})$ \\
    \texttt{db21} & $\mathcal{O}(nv)$ & $\mathcal{O}(nv^2)$ \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Explaining the performance benefits.} Let $n = |\mathcal{V}|$ be the
number of vertices in the Bayesian network, $d = \max_{X \in \mathcal{V}}
|\mathrm{pa}(X)|$ the maximum in-degree (i.e., number of parents), and $v =
\max_{X \in \mathcal{V}} |\im X|$ the maximum number of values per variable.
\Cref{tbl:asymptotes} shows how \texttt{db21} has both fewer variables and fewer
ADDs than any other encoding. However, note that these are upper bounds and most
encodings (including \texttt{db21}) can be smaller in certain situations (e.g.,
with binary random variables or when a CPT has repeating probabilities). We
equate clauses and ADDs (more specifically, factors of the function $\phi$ from
\cref{alg:encoding}) here because ADDMC interprets each clause of any WMC
encoding as a multiplicative factor of the ADD that represents the entire WMC
instance \cite{DBLP:conf/aaai/DudekPV20}. For literal-weight encodings, each
weight is also a factor, but that does not affect the asymptotic bounds in the
table.

>>>>

\begin{abstract}
  Weighted model counting (WMC) has emerged as the unifying inference mechanism
  across many (probabilistic) domains. Encoding an inference problem as an
  instance of WMC typically necessitates adding extra literals and clauses. This
  is partly so because the predominant definition of WMC assigns weights to
  models based on weights on literals, and this severely restricts what
  probability distributions can be represented. We develop a measure-theoretic
  perspective on WMC and propose a way to encode conditional weights on literals
  analogously to conditional probabilities. This representation can be as
  succinct as standard WMC with weights on literals but can also expand as
  needed to represent probability distributions with less structure. To
  demonstrate the performance benefits of conditional weights over the addition
  of extra literals, we develop a new WMC encoding for Bayesian networks and
  adapt a state-of-the-art WMC algorithm ADDMC to the new format. Our
  experiments show that the new encoding results in significantly faster
  inference on most benchmark instances. Furthermore, the same idea can be
  adapted to other WMC algorithms and other problem domains.
\end{abstract}

@@

% TODO: Feedback: our empirical evaluations show a x% improvement

Weighted model counting (WMC), i.e., an extension of model counting ($\#\SAT{}$)
that assigns a weight to every model \cite{DBLP:conf/aaai/SangBK05}, has emerged
as one of the most dominant and competitive approaches for handling inference
tasks in a wide range of formalisms including Bayesian networks
\cite{DBLP:conf/aaai/SangBK05,DBLP:books/daglib/0024906}, probabilistic
graphical models more generally \cite{DBLP:conf/ecsqaru/ChoiKD13}, and
probabilistic programs
\cite{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089}.
Over the last fifteen years, WMC has been extended and generalised in many ways,
e.g., to handle continuous probability distributions
\cite{DBLP:conf/ijcai/BellePB15}, first-order probabilistic theories
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16}, and infinite
domains \cite{DBLP:conf/aaai/Belle17}. Furthermore, by generalising the notion
of weights to an arbitrary semiring, a range of other problems are also captured
\cite{DBLP:journals/japll/KimmigBR17}. Exact WMC solvers typically rely on
either knowledge compilation
\cite{DBLP:conf/ijcai/OztokD15,DBLP:conf/ijcai/LagniezM17} or exhaustive DPLL
search \cite{DBLP:conf/aaai/SangBK05}, whereas approximate solvers work by
sampling \cite{DBLP:conf/aaai/ChakrabortyFMSV14} and performing local search
\cite{DBLP:conf/sat/WeiS05}.

@@

@@

%making it applicable to domains such as parsing, dynamic programming, constraint
%programming, databases, and networks.

@@

\paragraph{ADDs and their use in inference.} ADDs provide an efficient way to
manipulate functions from a Boolean algebra to any algebraic structure (most
commonly real numbers) \cite{DBLP:journals/fmsd/BaharFGHMPS97}. They have been
used to represent value functions in Markov decision processes (MDPs)
\cite{DBLP:conf/uai/HoeySHB99} and extended to handle first-order MDPs
\cite{DBLP:journals/ai/SannerB09} and discrete and continuous state MDPs
\cite{DBLP:conf/uai/SannerDB11}. The latter extension has also been used for
weighted model integration---an extension of WMC that can handle continuous
variables \cite{DBLP:conf/ijcai/KolbMSBK18}. For Bayesian networks, ADDS have
been used to represent each conditional probability table (CPT) as an ADD
\cite{DBLP:conf/icml/ZhaoMP15} and as a combination of ADDs and arithmetic
circuits \cite{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous
in this situation because of their ability to fully exploit context-specific
independence, i.e., the observable structure within a CPT that is not inherited
from the structure of the Bayesian network \cite{DBLP:conf/uai/BoutilierFGK96}.

@@

Note that in this case, we chose to interpret $\nu_b$ as $\Pr(b \mid a)$
while---with a different definition of $\nu_b$ that represents the joint
probability distribution $\Pr(a, b)$---$\nu_b$ by itself could induce $\mu$. In
general, however, factorising the full weight function into several smaller
functions often results in weight functions with smaller domains which leads to
increased efficiency and decreased memory usage \cite{DBLP:conf/aaai/DudekPV20}.
We can easily generalise this example further.

\begin{theorem}
  For any set $U$ such that $|U| \ge 2$, there exists a non-factorable measure
  $2^{2^{U}} \to \mathbb{R}_{\ge 0}$.
\end{theorem}

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\cite{DBLP:journals/ai/ChaviraD08}. We can use the measure-theoretic perspective
on WMC to show that this is always possible, however, as ensuing sections will
demonstrate, it often makes the inference task much harder in
practice.\footnote{The proofs of this and other theoretical results can be found
in the technical appendix.}

@@

In this section, we describe a way to encode Bayesian networks into WMC without
restricting oneself to factorable measures and thus having to add extra
variables. We will refer to it as \texttt{cw}.

A Bayesian network is a directed acyclic graph with random variables as vertices
that defines a probability distribution over them. Let $\mathcal{V}$ denote this
set of random variables. For any random variable $X \in \mathcal{V}$, let $\im
X$ denote its set of values and $\mathrm{pa}(X)$ its set of parents. The
full probability distribution is then equal to $\prod_{X \in \mathcal{V}} \Pr(X
\mid \mathrm{pa}(X))$ \cite{DBLP:books/daglib/0023820}. For discrete Bayesian
networks (and we only consider discrete networks in this paper), each factor of
this product can be represented by a CPT. See \cref{fig:example_bn} for an
example Bayesian network that we will refer to throughout this section. For
this network, $\mathcal{V} = \{ W, F, T \}$, $\mathrm{pa}(W) = \emptyset$,
$\mathrm{pa}(F) = \mathrm{pa}(T) = \{ W \}$, $\im W = \im F = \{0, 1 \}$, and
$\im T = \{ l, m, h \}$.

@@

For a binary random variable $X$, $\mathrm{CPT}_X$ is simply a sum of smaller
functions, one for each row of the CPT. If $X$ has more than two values, we also
multiply $\mathrm{CPT}_X$ by `clause' functions that restrict the value of
$\phi(T)$ to zero whenever $|\mathcal{E}(X) \cap T| \ne 1$. For the Bayesian
network in \cref{fig:example_bn}, we get:
\begin{align*}
  \mathrm{CPT_F} &= 0.6[\lambda_{F=1}] \cdot [\lambda_{W=1}] + 0.4[\lambda_{F=0}] \cdot [\lambda_{W=1}] \\
                 &+ 0.1[\lambda_{F=1}] \cdot [\lambda_{W=0}] + 0.9[\lambda_{F=0}] \cdot [\lambda_{W=0}], \\
  \mathrm{CPT_T} &= ([\lambda_{T=l}] + [\lambda_{T=m}] + [\lambda_{T=h}]) \\
                 &\cdot (\overline{[\lambda_{T=l}]} + \overline{[\lambda_{T=m}]}) \cdot (\overline{[\lambda_{T=l}]} + \overline{[\lambda_{T=h}]}) \\
                 &\cdot (\overline{[\lambda_{T=m}]} + \overline{[\lambda_{T=h}]}) \cdot \dots.
\end{align*}

@@

\Cref{alg:encoding} produces a function with a Boolean algebra as its domain.
This function can be represented by an ADD
\cite{DBLP:journals/fmsd/BaharFGHMPS97}. The core of ADDMC works by taking an
ADD $\psi\colon 2^{U} \to \mathbb{R}_{\ge 0}$ (expressed as a product of smaller
ADDs) and returning $(\exists_U\psi)(\emptyset)$
\cite{DBLP:conf/aaai/DudekPV20}. In this section, we prove that the function
$\phi$ produced by \cref{alg:encoding} can be used by ADDMC to correctly compute
any marginal probability of the Bayesian network that was encoded as
$\phi$.\footnote{Note that it can just as well compute \emph{any} probability
  expressed using the random variables in $\mathcal{V}$.} We begin with
\cref{lemma:cpt} which shows that any conditional weight function produces the
right answer when given a valid encoding of variable-value assignments relevant
to the CPT.

@@

For all encodings other than \texttt{cw}, we use their implementation in
Ace~3.0\footnote{\url{http://reasoning.cs.ucla.edu/ace/}} with
\texttt{-encodeOnly} and \texttt{-noEclause} flags. However, Ace was not used to
encode evidence, as preliminary experiments revealed that the evidence-encoding
implementation contains bugs that can lead to incorrect answers or a Java
exception being thrown on some instances of the data set (and the source code is
not publicly available). Instead, we simply list all the evidence as additional
clauses in the encoding.

@@

While encoding time itself was not measured because of different languages of
implementation, it is worth noting that \texttt{cw} is linear in the total
number of CPT rows and so is unlikely to be slower than, e.g., \texttt{cd06},
which relies on solving $\NP{}$-complete problems as part of the encoding process
\cite{DBLP:conf/sat/ChaviraD06}.

@@

\begin{table}
  \centering
  \caption{The numbers of instances solved by ADDMC with each encoding
    (uniquely, faster than with others, and in total) under the described time
    and memory constraints (out of 1216 instances)}%
  \begin{tabular}{lrrr}
    \toprule
    Encoding & Unique & Fastest & Total \\
    \midrule
    \texttt{cd05} & 2 & 3 & 372 \\
    \texttt{cd06} & 0 & 1 & 351 \\
    \texttt{cw} & 228 & 871 & 901 \\
    \texttt{d02} & 41 & 99 & 726 \\
    \texttt{sbk05} & 6 & 36 & 687 \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{figure*}
  \centering
  \input{scatter.tex}%
  \caption{ADDMC inference time using \texttt{cw} compared to \texttt{d02}
    (left) and \texttt{sbk05} (right) for each instance in all data sets.}
\end{figure*}

% TODO: write. then run through a grammar checker.
\paragraph{Observations.}
\begin{itemize}
\item The order of the encodings from best to worst is exactly the opposite of
  that in the previous literature.
\item Our encoding is particularly promising on instances of Grid networks.
  They are set up so that the entire Bayesian network is relevant to the
  query.
\item However, \texttt{cw} struggles with instances from Mastermind and
  Random Blocks domains (but so does \texttt{sbk05}). We conjecture that this
  is so either because the particular structure of these problems confuse the
  heuristics used by ADDMC or because these instances allow for significant
  simplifications that are exploited by \texttt{d02} but not \texttt{cw}.
\item An observation from the cumulative plot: \texttt{cw} solves the same
  number of instances in \SI{6.374}{\second} as \texttt{d02} does in
  \SI{1000}{\second}.
\end{itemize}

\begin{table}
  \centering
  \caption{Asymptotic upper bounds on the numbers of variables and clauses/ADDs
    for each encoding}
  \label{tbl:asymptotes}
  \begin{tabular}{lcc}
    \toprule
    Encoding(s) & Variables & Clauses/ADDs \\
    \midrule
    \texttt{cd05}, \texttt{cd06}, \texttt{sbk05} & $\mathcal{O}(nv^{d+1})$ & $\mathcal{O}(nv^{d+1})$ \\
    \texttt{cw} & $\mathcal{O}(nv)$ & $\mathcal{O}(nv^2)$ \\
    \texttt{d02} & $\mathcal{O}(nv^{d+1})$ & $\mathcal{O}(ndv^{d+1})$ \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Explaining the performance benefits.} Let $n = |\mathcal{V}|$ be the
number of vertices in the Bayesian network, $d = \max_{X \in \mathcal{V}}
|\mathrm{pa}(X)|$ the maximum in-degree (i.e., number of parents), and $v =
\max_{X \in \mathcal{V}} |\im X|$ the maximum number of values per variable.
\Cref{tbl:asymptotes} shows how \texttt{cw} has both fewer variables and fewer
ADDs than any other encoding. However, note that these are upper bounds and most
encodings (including \texttt{cw}) can be smaller in certain situations (e.g.,
with binary random variables or when a CPT has repeating probabilities). We
equate clauses and ADDs (more specifically, factors of the function $\phi$ from
\cref{alg:encoding}) here because ADDMC interprets each clause of any WMC
encoding as a multiplicative factor of the ADD that represents the entire WMC
instance \cite{DBLP:conf/aaai/DudekPV20}. For literal-weight encodings, each
weight is also a factor, but that does not affect the asymptotic bounds in the
table.
